--- 
title: "Analytical Portfolio"
author: "Valerie C. Valerio"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This website contains examples from Valerie's analytical portfolio."
---

# ABout this site

This website contains short _analytical examples_ from Valerie's portfolio. These examples might include code and analyses carried out as a PhD student at the University of Florida, as a Research Scientist at Surgo Ventures, and/or as side (aka passion) projects. If you have any questions, please don't hesitate to reach out (valerievalerioh@gmail.com). To see more of my work, please visit the Resources (Chapter \@ref(resources)).

The website includes text, figures and R code chunks. Code chunks will appear as follows:

```{r eval=FALSE}
# This is a comment in an R code chunk
# Comments help my colleagues (and Valerie from the future) understand what the code is supposed to do and why

## Install packages -----
install.packages("tidyverse") 
```



```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
